Hippocampal replay of experience at real-world speeds

### 2020-10-08 Article.tex

We find that the large majority of SWRs contain spatially coherent content; that is, the replay trajectories that are spatially concentrated at each moment in time and have no large jumps. Surprisingly, while the expected high speed, sequential replay events were identified, the most common category of events expressed representations that moved at much slower speeds consistent with real-world experiences, suggesting that high speed replay is not the most common manifestation of generative activity. These findings illustrate the power of state space models and provide a new understanding of the nature of hippocampal replay.

@@

We found that spatially coherent events did indeed have lower cumulative HPD sizes than events with fragmented dynamics. Figure \ref{4}A and D show two example SWRs that were classified as having stationary and continuous dynamics, respectively. The majority of the positions in the HPD at each time step in these SWRs is concentrated in a small portion of the track and the cumulative HPD size is low throughout the SWRs. In contrast, Figure \ref{4}C shows a SWR where the dynamics are fragmented and correspondingly, the HPD is much more spatially diffuse and the cumulative HPD size is much higher. The HPD also provides insights into the moment-by-moment structure of each event, which can change over the time course of a SWR. An example of this change is shown in Figure \ref{4}B, where the cumulative HPD size is small for most of the SWR until the end, at which point the uncertainty of position becomes much higher, reflecting a transition from a spatially coherent to a spatially incoherent representation. Overall, when we examined the average cumulative HPD size for each SWR, grouped by dynamic, we found a clear bimodal split between spatially coherent dynamics and spatially incoherent dynamics (Figure \ref{4}E). For the spatially coherent dynamics, the cumulative HPD size was much lower than the spatially incoherent dynamics (median 24 cm, spatially coherent vs. median 238 cm, spatially incoherent, p=2.2e-16, one-sided Mann-Whitney-U). Additionally, the HPD size for the unclassified SWRs was also We note here that while the size of the HPD will be influenced by the dynamic estimated from the data, it remains the case that a continuous or stationary dynamic can correspond to a large HPD region (see outliers in Figure \ref{4}E), indicating less certainty for those specific events.

@@

\subsection*{Many replay trajectories have slower movement speeds}
\begin{figure*}%[tbhp]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure5/Figure5_v4}
\caption{
Prevalence of classifications. \textbf{(A)} UpSet plot \cite{LexUpSetVisualizationIntersecting2014}---which is similar to a Venn diagram with more than 3 sets---of the most common sets of classifications within each SWR. Each row represents a classification and each column represents a set of classifications, where filled-in black dots with an edge between the dots indicates that multiple classifications are present in the SWR (at temporally distinct times). The sets are ordered by how often they occur as indicated by the bar plot above each category. The total percentage of each classification is indicated by the rotated bar plot on the left. \textbf{(B)} Duration of each dynamic category within a SWR. The box shows the interquartile range (25-75\%) of the data and the whiskers are 1.5 times the interquartile range. Points outside of this are labeled as diamonds. \textbf{(C)} Average distance between the latent position and the animal's actual position for each classification within the SWR. \textbf{(D)} Average speed of the classification within the SWR (excluding classifications < 20 ms).\textbf{(E)} Kernel density estimate of the position of stationary trajectories on the W-track at least 30 cm away from the animal's position. The shaded region represents the density estimate while the black ticks represent the observed non-local stationary positions. \textbf{(F)} Average tetrode multiunit spike rates for each dynamic category within each SWR (excluding classifications < 20 ms).
}
\label{5}
\end{figure*}

@@

These results challenge the long-held notion that hippocampal replay trajectories necessarily proceed at many times the speed of the animal \cite{NadasdyReplayTimeCompression1999, LeeMemorySequentialExperience2002, DavidsonHippocampalReplayExtended2009}. Instead, there is a much richer variety of speeds the animal can use to retrieve information about prior experience or generate possible future events, including slower moving representations that activate a spatially compact set of nearby locations and stationary representations of single locations. Interestingly, a previous paper \cite{FarooqEmergencepreconfiguredplastic2019} reported that these stationary representations were most prevalent in juvenile rats and decreased to "chance" levels when the rats reached adulthood. Here, we show that they do not disappear, but are a very common occurrence with adult rats. We believe that the difference in these findings is a result of challenge of defining "chance" levels of occurrence of these events. We also note that we chose to include a stationary distribution to explicitly evaluate the prevalence of stationary events in adult animals, but in principle a model with just continuous and fragmented dynamics could capture these events as well, as the continuous dynamic includes movement speeds close to 0 m/s.

>>>>

We find that the large majority of SWRs contain spatially coherent content; that is, the replay trajectories that are spatially concentrated at each moment in time and have no large jumps. Surprisingly, while the expected high speed, sequential replay events were identified, the most common category of events expressed representations that moved at much slower speeds, more consistent with real-world experiences, suggesting that high speed replay is not the most common manifestation of generative activity. These findings illustrate the power of state space models and provide a new understanding of the nature of hippocampal replay.

@@

We found that spatially coherent events did indeed have lower cumulative HPD sizes than events with fragmented dynamics. Figure \ref{4}A and D show two example SWRs that were classified as having stationary and continuous dynamics, respectively. The most probable positions (as estimated by the model) at each time step in these SWRs is concentrated in a small portion of the track and correspondingly, the cumulative HPD size is small throughout the SWRs. In contrast, Figure \ref{4}C shows a SWR where the dynamics are fragmented and correspondingly, the most probable positions are more spatially diffuse and the cumulative HPD size is much larger. The HPD size also provides insights into the moment-by-moment structure of each event, which can change over the time course of a SWR. An example of this change is shown in Figure \ref{4}B, where the cumulative HPD size is small for most of the SWR until the end, at which point the uncertainty of position becomes much higher, reflecting a transition from a spatially coherent to a spatially incoherent representation. Overall, when we examined the average cumulative HPD size for each SWR, grouped by dynamic, we found a clear bimodal split between spatially coherent dynamics and spatially incoherent dynamics (Figure \ref{4}E). For the spatially coherent dynamics, the average cumulative HPD size for each SWR was much smaller than the spatially incoherent dynamics (median 24 cm, spatially coherent vs. median 238 cm, spatially incoherent, p=2.2e-16, one-sided Mann-Whitney-U). Additionally, the HPD size for the unclassified SWRs was similarly large. We note here that while the size of the HPD will be influenced by the dynamic estimated from the data, it remains the case that a continuous or stationary dynamic can correspond to a large HPD region (see outliers in Figure \ref{4}E), indicating less certainty for those specific events.  To show this, as before, we resampled position with replacement for two recording sessions, shuffling the relationship between spiking and spatial information and then fitting the encoding model. The shuffled decodes had much larger HPD size than the real data (epoch \#1: 55 cm vs. 229 cm, p=0.02, real vs. shuffled; epoch \#2: 72 cm vs. 231 cm, p=0.02, real vs. shuffled; one-sided Mann-Whitney-U, \ref{fig:Figure4-Figure supplement 1}).

@@

\subsection*{Many replay trajectories have slower movement speeds}
\begin{figure*}%[tbhp]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure5/Figure5_v4}
\caption{
Prevalence of classifications. \textbf{(A)} UpSet plot \cite{LexUpSetVisualizationIntersecting2014}---which is similar to a Venn diagram with more than 3 sets---of the most common sets of classifications within each SWR. Each row represents a classification and each column represents a set of classifications, where filled-in black dots with an edge between the dots indicates that multiple classifications are present in the SWR (at temporally distinct times). The sets are ordered by how often they occur as indicated by the bar plot above each category. The total percentage of each classification is indicated by the rotated bar plot on the left. \textbf{(B)} Duration of each dynamic category within a SWR. The box shows the interquartile range (25-75\%) of the data and the whiskers are 1.5 times the interquartile range. Points outside of this are labeled as diamonds. \textbf{(C)} Average distance between the latent position and the animal's actual position for each classification within the SWR. \textbf{(D)} Average speed of the classification within the SWR (excluding classifications with durations < 20 ms). Note that these speeds are calculated using the most probable position (MAP estimate) which can be noisy when the probability of position is flat or multimodal.\textbf{(E)} Kernel density estimate of the position of stationary trajectories on the W-track at least 30 cm away from the animal's position. The shaded region represents the density estimate while the black ticks represent the observed non-local stationary positions. \textbf{(F)} Average tetrode multiunit spike rates for each dynamic category within each SWR (excluding classifications < 20 ms).
}
\label{5}
\end{figure*}

@@

These results challenge the long-held notion that hippocampal replay trajectories necessarily proceed at many times the speed of the animal \cite{NadasdyReplayTimeCompression1999, LeeMemorySequentialExperience2002, DavidsonHippocampalReplayExtended2009}. Instead, there is a much richer variety of speeds the animal can use to retrieve information about prior experience or generate possible future events, including slower moving representations that activate a spatially compact set of nearby locations and stationary representations of single locations. These representation do not move far in the span of a SWR. Interestingly, a previous paper \cite{FarooqEmergencepreconfiguredplastic2019} reported that these stationary representations were most prevalent in juvenile rats and decreased to "chance" levels when the rats reached adulthood. Here, we show that they do not disappear, but are a very common occurrence with adult rats. We believe that the difference in these findings is a result of defining "chance" levels of occurrence of these events, using large time bins, and using a restrictive decoding model. We also note that we chose to include a stationary distribution to explicitly evaluate the prevalence of stationary events in adult animals, but in principle a model with just continuous and fragmented dynamics could capture these events as well, as the continuous dynamic includes movement speeds close to 0 m/s.
